\Title

### 2013-02-12 supplementary.tex
Fig.1 has p10; Fig S3 -> S3/S4; more details on epitopes; S2 larger

\newpage
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Synonymous diversity across the HIV genome}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=\linewidth]{conservation_codons_genome}
\caption{Synonymous diversity across the HIV genome, as quantified by the
normalized codon entropy among sequences coding for the
consensus amino-acid.
The synonymous diversity peaks at the variable regions in {\it env} and is reduced in regions under purifying selection (RRE
hairpin, second {\it tat}/{\it rev} exons). Since the variable regions are under
positive selection at the protein level, synonymous hitchhiking is stronger
there, and even deleterious synonymous mutations can be brought to high
frequency. This gives to our signal in the fixation probability, Figs. 2A and
2B of the main text.
The normalized codon entropy is calculated as follows (see the script
for the full algorithm): (i) from a subtype B multiple sequence alignment
(MSA) from the LANL website (filtered sequences only, version 2011)~\cite{LANL2012},
we calculate the consensus amino acid at each position in the HIV genome; (ii) we count
how often each codon coding for the consensus amino acid appears in the MSA;
(iii) at each amino acid position, we divide by the number of sequences in the
MSA that had the consensus amino acid at that position, obtaining {\it codon
frequencies} $\nu_c$; (iv) we calculate the codon entropy from each position as:
$S := - \sum_{c} \nu_c \log \nu_c$, where $c$ runs over codons that code for
the consensus amino acid at this site; (v) we divide by the maximal codon
entropy of that amino acid (e.g. $\log 2$ for twofold degenerate codons).
All parts of {\it env} that are part of a different gene (signaling peptide,
second {\it rev} exon) have been excluded from our main analysis, to avoid
contamination about synonymity.}
\label{fig:syndiv_genome}
\end{center}
\end{figure}
\newpage
% 
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section{Nonsynonymous changes outside of variable regions are deleterious}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \begin{figure}[h]
% \begin{center}
% \includegraphics[width=0.8\linewidth]{synmut_conservation_4fold_synnonsyn}
% \caption{Cumulative distribution of synonymous and nonsynonymous diversity in
% {\it gag} in the LANL reference panel (filtered
% sequences only, version 2011)~\cite{LANL2012}. Sites such that the consensus codon
% has three synonymous and six nonsynonymous single mutants were used, and the number of observed mutants
% of a certain type over the number of possible mutants is plotted. Nonsynonymous
% changes are observed less often, {\it ergo} are more conserved, than
% synonymous changes. It can be therefore assumed, as mentioned in the main text,
% that non-escape nonsynonymous changes involve a large fitness cost.}
% \label{fig:synnonsyncons}
% \end{center}
% \end{figure}
% \newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Time dependent selection}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.8\linewidth]{simulations_graduallyepitopesandtimeselec}
\caption{Time-dependent selection (the escape mutant is recognized after some
months from its appearance) and competition between escape mutations in the same
epitope both reduce fixation of nonsynonymous mutations. A ``na\"ive'' computer
simulation shows a depression in synonymous fixation probability, as observed in
patients, but the nonsynonymous fixation probability is much higher than the
(neutral) diagonal: a nonsynonymous mutation that reaches a frequency around
0.2 is already very likely to fix. In data from infected patients, however,
nonsynonymous mutations show a neutral-like curve,
$P_\text{fix}(\nu) \approx \nu$. In the other curves, we show that the
nonsynonymous fixation probability can be reduced by either time-dependent
selection or competition between equivalent escapes, without affecting much the
synonymous fixation probability itself.
In the curves denoted with ``tds'' (time-dependent selection), the fitness advantage of
an escape mutaition can vanish at each time point with a probability density
\[ P_\text{recognized}(t) = c \cdot \nu(t), \]
where $c$ is a constant coefficient shown in the legend that encodes the overall
efficiency of the host immune system, and $\nu(t)$ is the frequency of the
escape allele at time $t$.
In the curves denoted with ``mut/epi'' we test the
competition scenario instead, with 3 or 6 escape mutations within the same
epitope of length 9 nucleotides. The fitness landscape of each such epitope
include negative epistatic terms, so that the joint presence of more than one
escape mutation is not any more beneficial for the virus than a single mutation.
In this latter scenario, many escape mutations start to sweep on different
backgrounds within the viral population, but eventually compete and only one of
them if fixed.}
\label{fig:tds_wec}
\end{center}
\end{figure}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Within epitope competition}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.8\linewidth]{simulations_graduallyepitopesandtimeselec}
\caption{Time-dependent selection (the escape mutant is recognized after some
months from its appearance) and competition between escape mutations in the same
epitope both reduce fixation of nonsynonymous mutations. A ``na\"ive'' computer
simulation shows a depression in synonymous fixation probability, as observed in
patients, but the nonsynonymous fixation probability is much higher than the
(neutral) diagonal: a nonsynonymous mutation that reaches a frequency around
0.2 is already very likely to fix. In data from infected patients, however,
nonsynonymous mutations show a neutral-like curve,
$P_\text{fix}(\nu) \approx \nu$. In the other curves, we show that the
nonsynonymous fixation probability can be reduced by either time-dependent
selection or competition between equivalent escapes, without affecting much the
synonymous fixation probability itself.
In the curves denoted with ``tds'' (time-dependent selection), the fitness advantage of
an escape mutaition can vanish at each time point with a probability density
\[ P_\text{recognized}(t) = c \cdot \nu(t), \]
where $c$ is a constant coefficient shown in the legend that encodes the overall
efficiency of the host immune system, and $\nu(t)$ is the frequency of the
escape allele at time $t$.
In the curves denoted with ``mut/epi'' we test the
competition scenario instead, with 3 or 6 escape mutations within the same
epitope of length 9 nucleotides. The fitness landscape of each such epitope
include negative epistatic terms, so that the joint presence of more than one
escape mutation is not any more beneficial for the virus than a single mutation.
In this latter scenario, many escape mutations start to sweep on different
backgrounds within the viral population, but eventually compete and only one of
them if fixed.}
\label{fig:tds_wec}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{natbib}
\bibliography{bib}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\newpage
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Synonymous diversity across the HIV genome}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=\linewidth]{conservation_codons_genome}
\caption{
Synonymous diversity across the HIV genome, as quantified by the normalized
codon entropy among sequences coding for the consensus amino-acid. The
synonymous diversity peaks at the variable regions in {\it env} and is reduced
in regions under purifying selection (RRE hairpin, second {\it tat}/{\it rev}
exons). Since the variable regions are under positive selection at the protein
level, synonymous hitchhiking is stronger there, and even deleterious synonymous
mutations can be brought to high frequency. This gives to our signal in the
fixation probability, Figs. 2A and 2B of the main text. The normalized codon
entropy is calculated as follows (see the script for the full algorithm): (i)
from a subtype B multiple sequence alignment (MSA) from the LANL website
(filtered sequences only, version 2011)~\cite{LANL2012}, we calculate the
consensus amino acid at each position in the HIV genome; (ii) we count how often
each codon coding for the consensus amino acid appears in the MSA; (iii) at each
amino acid position, we divide by the number of sequences in the MSA that had
the consensus amino acid at that position, obtaining {\it codon frequencies}
$\nu_c$; (iv) we calculate the codon entropy from each position as: $S := -
\sum_{c} \nu_c \log \nu_c$, where $c$ runs over codons that code for the
consensus amino acid at this site; (v) we divide by the maximal codon entropy of
that amino acid (e.g. $\log 2$ for twofold degenerate codons). All parts of
{\it env} that are part of a different gene (signaling peptide, second {\it rev}
exon) have been excluded from our main analysis, to avoid contamination about
synonymity. Note that all gap-rich columns of the MSA are stripped from this
figure, hence genes such as {\it env} might appear shorter than they actually
are.
}
\label{fig:syndiv_genome}
\end{center}
\end{figure}
\newpage
% 
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section{Nonsynonymous changes outside of variable regions are deleterious}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \begin{figure}[h]
% \begin{center}
% \includegraphics[width=0.8\linewidth]{synmut_conservation_4fold_synnonsyn}
% \caption{Cumulative distribution of synonymous and nonsynonymous diversity in
% {\it gag} in the LANL reference panel (filtered
% sequences only, version 2011)~\cite{LANL2012}. Sites such that the consensus codon
% has three synonymous and six nonsynonymous single mutants were used, and the number of observed mutants
% of a certain type over the number of possible mutants is plotted. Nonsynonymous
% changes are observed less often, {\it ergo} are more conserved, than
% synonymous changes. It can be therefore assumed, as mentioned in the main text,
% that non-escape nonsynonymous changes involve a large fitness cost.}
% \label{fig:synnonsyncons}
% \end{center}
% \end{figure}
% \newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Time-dependent selection}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\linewidth]{simulations_gradually_timeselec_deadagain}
\caption{
Time-dependent selection (the escape mutant is recognized after some months from
its appearance) reduces fixation of nonsynonymous
mutations. A ``na\"ive'' computer simulation shows a depression in synonymous
fixation probability, as observed in patients, but the nonsynonymous fixation
probability is much higher than the (neutral) diagonal: a nonsynonymous mutation
that reaches a frequency around 0.2 is already very likely to fix. In data from
infected patients, however, nonsynonymous mutations show a neutral-like curve,
$P_\text{fix}(\nu) \approx \nu$. In the other curves, we show that the
nonsynonymous fixation probability can be reduced by time-dependent selection
without affecting much the synonymous fixation probability itself. The fitness
advantage of an escape mutaition can vanish at each time point with a
probability density \[ P_\text{recognized}(t) = c \cdot \nu(t), \] where $c$ is
a constant coefficient shown in the legend that encodes the overall efficiency
of the host immune system, and $\nu(t)$ is the frequency of the escape allele at
time $t$. After recognition, the former escape mutation has the same fitness
cost as before the creation of the epitope, hence it is quickly lost. Note that,
although the curves shown here show almost no reduction in fixation probability
up to $\nu \approx 0.15$, this effect is not robust and depends on the exact form of
$ P_\text{recognized}(t)$. The other parameters for the shown simulations are
the following: deleterious effect $10^{-3}$, average escape rate $0.016$,
fraction of deleterious synonymous mutations $0.986$, rate of new epitopes
$0.0014$ per generation.
}
\label{fig:tds}
\end{center}
\end{figure}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Within-epitope competition}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\linewidth]{simulations_gradually_epitopes}
\caption{
Competition between escape mutations in the same epitope also reduces fixation of
nonsynonymous mutations. Compared to a ``na\"ive'' model,
simulations with 3 or 6 escape mutations competing within the same epitope show
reduced fixation probabilities of nonsynonymous changes, whereas the synonymous
changes behave similarly in all cases. The epitope length is 9 nucleotides. The
fitness landscape of each such epitope include negative epistatic terms, so that
the joint presence of more than one escape mutation is not any more beneficial
for the virus than a single mutation. In detail, each site has two alleles, $\pm
1$, where $-1$ is the ancestral one and $+1$ the dervied one; the fitness
coefficient of a $k$-uple of sites within the epitope is $f_k = \pm
\eta_\epsilon / 2^{L-1}$, where $\eta_\epsilon$ is a random variable for the
escape rate with average value $\epsilon$ (see methods), $L$ is the number of
competing escapes in the epitope, and the positive (resp. negative) sign is for
odd (resp. even) $k$. In this evolutionary scenario, many escape mutations start
to sweep on different backgrounds within the viral population, but eventually
compete and only one of them if fixed. The other parameters for the shown simulations are
the following: deleterious effect $10^{-3}$, average escape rate $0.016$,
fraction of deleterious synonymous mutations $0.986$, rate of new epitopes
$0.0014$ per generation.
}
\label{fig:wec}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{natbib}
\bibliography{bib}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-02-12 synmut.tex
Fig.1 has p10; Fig S3 -> S3/S4; more details on epitopes; S2 larger

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency.  A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e.~with
probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{aftsyn},
this is a simple consequence of the fact that
(i) exactly one individual in the current population will be
the common ancestor of the entire future population at a particular locus and
(ii) this ancestor has a probability $\nu$ of carrying the mutation (assuming
the neutral mutation is not preferentially associated with genomes of high or
low fitness).
Deleterious or beneficial mutations fix less or
more often than neutral ones, respectively. \FIG{aft} shows 
the time course the frequencies of all synonymous and nonsynonymous mutations
observed \env, C2-V5, in patient p1~\citep{shankarappa_consistent_1999},
respectively. Despite many synonymous mutations reaching high frequency, 
few fix (panel~\ref{fig:aftsyn}); in constrast, many nonsynonymous mutations fix
(panel~\ref{fig:aftnonsyn}).

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p1.pdf}
\label{fig:aftsyn}}\\
 \subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p1.pdf}
\label{fig:aftnonsyn}}
\caption{\comment{I would like p2 or p10 better here} Time series of frequencies
of synonymous (A) and non-synonymous (B) derived alleles in \env, C2-V5, from patient
1~\cite{shankarappa_consistent_1999}.
While many nonsynonymous mutations  fix, few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the C2-V5 region
(blue to red). Inset: the fixation probability $P_\text{fix}(\nu)$ of a neutral mutation
is simply the likelihood that the future common ancestor is currently carrying
it, i.e., its frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

@@

To get a better idea of the range of parameters that are compatible with the
observations and our interpretation, we performed computer simulations of
evolving viral populations assuming a mix of positive and purifying selection
and rare recombination.
For this purpose, we use the simulation package FFPopSim, which includes a
module dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}. 
For each simulation run, we specify the deleterious effect of synonymous
mutations, the fraction of synonymous mutations that are deleterious, the
escape rate of adaptive non-synonymous mutations and the frequency of new
escapes. Note that the escape rate is sum of two factors: (i) the beneficial
effect due to the ability to evade the immune system minus (ii) the fitness
cost of the mutation in terms of structure, stability, etc. Net escape rates
in chronic infections have been estimated to be on the order of 
$s_a = 0.01$ per day~\citep{neher_recombination_2010,Asquith:2006p28003}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Fixation probability and secondary structure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For the estimate of times to fixation/extinction, polymorphisms were
binned by frequency and the time to first reaching either fixation or
extinction was stored. The fixation probability was determined as the long-time
limit of the resulting curves. Mutations that reached high frequency but neither
fixed nor got extinct were classified as ``floating'' and not analyzed further.

@@

For the models with competition within epitopes, a complex epistatic fitness
landscape was designed such that each single mutant is sufficient for full
escape. \comment{need more}
To model recognition of escape variants by the immune system that is catching
up, the beneficial effect of an escape mutation was set to zero with a
probability per generation proportional to the frequency of the escape variant.

@@

The areas below or above the neutral fixation probability (diagonal line)
were estimated from the binned fixation probabilities using linear
interpolation between the bin centers. This measure is sufficiently precise for
our purposes. \comment{what happens with empty bins??}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency.  A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e.~with
probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{aftsyn},
this is a simple consequence of the fact that
(i) exactly one individual in the current population will be
the common ancestor of the entire future population at a particular locus and
(ii) this ancestor has a probability $\nu$ of carrying the mutation (assuming
the neutral mutation is not preferentially associated with genomes of high or
low fitness).
Deleterious or beneficial mutations fix less or
more often than neutral ones, respectively. \FIG{aft} shows 
the time course the frequencies of all synonymous and nonsynonymous mutations
observed \env, C2-V5, in patient p10~\citep{shankarappa_consistent_1999},
respectively. Despite many synonymous mutations reaching high frequency, 
few fix (panel~\ref{fig:aftsyn}); in constrast, many nonsynonymous mutations fix
(panel~\ref{fig:aftnonsyn}).

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
\label{fig:aftsyn}}\\
 \subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
\label{fig:aftnonsyn}}
\caption{Time series of frequencies
of synonymous (A) and non-synonymous (B) derived alleles in \env, C2-V5, from patient
1~\cite{shankarappa_consistent_1999}.
While many nonsynonymous mutations  fix, few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the C2-V5 region
(blue to red). Inset: the fixation probability $P_\text{fix}(\nu)$ of a neutral mutation
is simply the likelihood that the future common ancestor is currently carrying
it, i.e., its frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

@@

To get a better idea of the range of parameters that are compatible with the
observations and our interpretation, we performed computer simulations of
evolving viral populations assuming a mix of positive and purifying selection
and rare recombination.
For this purpose, we use the simulation package FFPopSim, which includes a
module dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}. 
For each simulation run, we specify the deleterious effect of synonymous
mutations, the fraction of synonymous mutations that are deleterious, the
escape rate of adaptive non-synonymous mutations and the frequency of new
escapes. Note that the escape rate is sum of two factors: (i) the beneficial
effect due to the ability to evade the immune system minus (ii) the fitness
cost of the mutation in terms of structure, stability, etc. Net escape rates
in chronic infections have been estimated to be on the order of 
$\epsilon = 0.01$ per day~\citep{neher_recombination_2010,Asquith:2006p28003}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Fixation probability and secondary structure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For the estimate of times to fixation/extinction, polymorphisms were
binned by frequency and the time to first reaching either fixation or
extinction was stored. The fixation probability was determined as the long-time
limit of the resulting curves. Mutations that reached high frequency but neither
fixed nor got extinct were classified as ``floating'', with one exception: if they first
reached high frequencies within 3 years of the last time point, we assumed
they had not had sufficient time to settle, so we discarded them.

@@

For the models with competition within epitopes, a complex epistatic fitness
landscape was designed such that each single mutant is sufficient for full
escape. In particular, each mutation had a linear effect equal to the escape,
but a negative epistatic effect of the same magnitude between each pair of sites
was included. Higher order terms compensated each other to make sure that not
only double, but all k-mutants with $k \geq 1$ had the same fitness. To model
recognition of escape variants by the immune system that is catching up, the
beneficial effect of an escape mutation was set to zero with a probability per
generation proportional to the frequency of the escape variant.

@@

The areas below or above the neutral fixation probability (diagonal line)
were estimated from the binned fixation probabilities using linear
interpolation between the bin centers. This measure is sufficiently precise for
our purposes. In 10 runs out of 2400, the highest frequency bin was empty so the
fixation probability could not be calculated; those runs were excluded from 
\FIG{simsfig}.
